---
title: "Data 607 HW2 SQL & R"
author: "Sufian"
date: "9/6/2019"
output: html_document
---

Instructions for Prof to run this code:

- Create a MYSQL Database called "projet-final"

- Run Hw2_v1 SQL script

- Run R Script

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

## Assignment 2 - SQL & R 

Creating database in SQL and importing into R and displaying the results in R

```{r, echo=F }
library("DT")
library(dplyr)
library(plotrix)
library(data.table)
library(RMySQL)
library(DBI)
library(rstudioapi)
require (ggplot2)

# connecting to projet-final database in MYSQL

# This line do not ask for user and password
con <- dbConnect(MySQL(), user ="root", password = "Chicke0n", port = 3306, dbname = "projet-final",host = "localhost")

# This line ask for user name and password
#con <- dbConnect(MySQL(), user = showPrompt('Username',"Username",default="root"), password = #askForPassword(), port = 3306, dbname = "projet-final",host = "localhost")

#query my reviews & videos tables from projet-final database

ret1 <- dbGetQuery(con,"select * FROM Videos")

ret2 <- dbGetQuery(con,"select * FROM Reviewer")


# show the tables
#movies table
datatable ( ret1 )

#reviewer table
datatable ( ret2 )

# check data structure of ret2
str(ret2)

# Converting NA to integer zero, to account for NULL Values in SQL Database
ret2[is.na(ret2)] <- 0

ret2

#bar plot of Total raw ratings
ggplot(data=ret2, aes(x=Name_of_Reviewer, y=Rating)) + geom_bar(stat="identity" )+
    geom_bar(colour="black", fill="#DD8888", width=.8, stat="identity" ) + 
    guides(fill=FALSE) +
    xlab("Participants") + ylab("Raw Ratings") +
    ggtitle("Sum of All Raw Ratings per 6 participants")

# Adding mean rating for each movie in the datatable

setDT(ret2)[, Mean_Rating := round(mean(Rating),digits = 1), by = Movie_Name]
datatable ( ret2)

# Get Unique Rows for Each movie with Rating and selecting only Movie_Name and Mean rating
dt_uniqe <- ret2[!duplicated(ret2[,'Movie_Name']),c(2,7)]
datatable ( dt_uniqe  )


#bar plot of Mean ratings of all Movies
ggplot(data=dt_uniqe, aes(x=Movie_Name, y=Mean_Rating)) + geom_bar(stat="identity")+
    geom_bar(colour="black", fill="#DD8888", width=.8, stat="identity") + 
    guides(fill=FALSE) +
    xlab("Movies") + ylab("Mean Ratings") +
    ggtitle("Mean Ratings by Movies")


#Creating Dataframe with all the movies and their mean rating and movie types
df <- merge(dt_uniqe, ret1, by="Movie_Name")
df


```
Summary Results:

First table - The movie table

2nd table - Reviewer table with original NA 

3rd table - Reviewer table with NA replaced by zero

Observation #1

Estimating the participation rate (how actively they were involved) by sum of raw ratings:

(a) Participants Steffi & Tom gave the most ratings; Adding up all their ratings per movie. This means these 2 participants was actively involved in the survey by providing scores 

(b) Participant Apu was on the other extreme end followed closely by Jeff;  these 2 participants either did not watch the movies or did was not actively participating in the survey at all

Observation #2

Estimating which movies scored best or worst via their average rating scores:

(a) The movies "something about Mary", "The International" and "Wonder woman" scored the highest respectively

(b) The movies "Lion King" was the worst rated followed by distance second "The souvenir"


links to RPubs:
http://rpubs.com/ssufian/525717

link to Github:

https://github.com/ssufian/Data_607_HW2_SQL_R

