
title: "Dat 607 HW 9 - Web APIs"

author: "Sufian"

date: "10/22/2019"

output: html_document


Rpub links:

http://rpubs.com/ssufian/543492


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Statement

- The New York Times web site provides a rich set of APIs, as described here: http://developer.nytimes.com/docs

- You’ll need to start by signing up for an API key.

- Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and

transform it to an R dataframe.

```{r message=FALSE, warning=FALSE}
#Load libraries
library(httr)
library(jsonlite)
library(tidyr)
library(lubridate)
library(rvest)
library(dplyr)
library(ggplot2)
library(tidyverse)
library("tm")
library("SnowballC")
library("wordcloud")
library("RColorBrewer")
```

## Testing the GET function to see if it works

```{r}
URL <- 'https://api.nytimes.com/svc/mostpopular/v2/viewed/30.json?api-key='

name <- "MfbAm3jsAPQ0UZ3kAxRGZ8SdbxcDVWQD"

myurl <- (paste0(URL, name, sep=""))

r <- GET(myurl)


r
```

## Error Check

```{r}
# check for error (TRUE if above 400)
http_error(r)

```

## Using Jasonlite to get articles from NY Times & Transforming into Dataframe

- This first pull was to get ALL the latest articles (any) for the last 23 days:  Oct 01, 2019 to Oct 23, 2019

```{r}
r2 <- fromJSON(myurl, flatten=TRUE) %>% data.frame()

head(r2)

```

## Peeking into the columns

```{r}
# Take a look at the columns
colnames(r2)
```

## Check how many articles

```{r}
# The search returned 20 articles with 22 columns bc each page/request has a max of 10 articles
dim(r2) 
```

## Make into a dataframe minus all the irrelavent columns

```{r}
# take only columns that is relevant from the data from NY times
final <- tibble("News_Source"=r2$results.source, "Title"=r2$results.title,"Authors"=r2$results.byline ,"News_type"=r2$results.type,"News_url"=r2$results.url,"News_abstract"=r2$results.abstract
               , "News_section"=r2$results.section)

head(final)
```

## Check news sources (which ones)

- Most news were from articles by 50%

```{r}
# Visualize coverage of articles by news type 
final %>% 
  group_by(News_type) %>%
  summarize(count=n()) %>%
  mutate(percent = (count / sum(count))*100) %>%
  ggplot() +
  geom_bar(aes(y=percent, x=News_type, fill=News_type), stat = "identity") + coord_flip()
```

##  WHich sections has most of the articles

- Most news are from the US news section

```{r}
# Visualize coverage of articles covered by sections
final %>% 
  group_by(News_section) %>%
  summarize(count=n()) %>%
  mutate(percent = (count / sum(count))*100) %>%
  ggplot() +
  geom_bar(aes(y=percent, x=News_section, fill=News_section), stat = "identity") + coord_flip()
```
```{r}
url_impeachment<- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=impeachment&api-key=",name, sep="")
impeachment1 <- GET(url_impeachment, accept_json())

```

## Lets try to see if the "Impeachment story" is a big deal?

- Using the search and the the any article url but with the same API key


```{r}

impeachment1 <- fromJSON(url_impeachment, flatten=TRUE) %>% data.frame()



# Take a look at the columns
colnames(impeachment1)

```

##  Repeat the same process again to see the dimensions

```{r}
dim(impeachment1) 
```

##  Taking just the last 7 days or so to see how many articles were related to the impeachment story

```{r}
# Set some parameters to grab all the hits by identifying a date range and max page # to loop through
term <- "impeachment" 
begin_date <- "20191001" # YYYYMMDD
end_date <- "20191023"
```

## Piecing the url together for the API call

```{r}
# Concatenate pieces of the url for the api call
baseurl <- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=",term,
                  "&begin_date=",begin_date,"&end_date=",end_date,
                  "&facet_filter=true&api-key=",name, sep="")
```


##  Determine the max no. of pages


```{r}
# Identify the # of hits to calculate the max pages 
initialQuery <- fromJSON(baseurl)
print(initialQuery$response$meta$hits[1]) # returns the total # of hits
```

## Looking at max pages found
`
```{r}
maxPages <- ceiling((initialQuery$response$meta$hits[1] / 10) -1) # reduce by 1 because loop starts with page 0
print(maxPages) # 88 is the max pages
```

## I had to limit the loop to only 10 pages so the connection would not timed out for illustratative purposes
`
```{r}
# Loop through all pages to get all the hits
pages <- list()
for(i in 0:10){
  nytSearch <- fromJSON(paste0(baseurl, "&page=", i), flatten = TRUE) %>% data.frame() 
  message("Retrieving page ", i)
  pages[[i+1]] <- nytSearch 
  Sys.sleep(9) # because there are 3 previous calls in under a min and the api call limit is 10/min 
}
```
```{r}
# Row bind the page results into a big a dataframe
impeachment_search <- rbind_pages(pages)

# Take a peek at 2 informative columns. 
head(impeachment_search , n=10)[c('response.docs.web_url', 'response.docs.snippet')]  

```

##  The impeachment stories/articles are mostly from News type; the impeachement coverages are virtually in News

```{r}
impeachment_search  %>% 
  group_by(response.docs.type_of_material) %>%
  summarize(count=n()) %>%
  mutate(percent = (count / sum(count))*100) %>%
  ggplot() +
  geom_bar(aes(y=percent, x=response.docs.type_of_material, fill=response.docs.type_of_material), stat = "identity") + coord_flip()
```

```{r}
# Visualize coverage of dreamers by section
impeachment_search  %>% 
  group_by(response.docs.section_name) %>%
  summarize(count=n()) %>%
  mutate(percent = (count / sum(count))*100) %>%
  ggplot() +
  geom_bar(aes(y=percent, x=response.docs.section_name, fill=response.docs.section_name), stat = "identity") + coord_flip()
```


##  The impeachment stories/articles are mostly from US News Sections.  Looks like the world does not really care

about the impeachment stories but the US


