---
title: "Data 607 Week 12 Assignment - NoSQL Migration"
author: "Sufian"
date: "11/08/2019"
output: html_document
---

## links to RPubs:

http://rpubs.com/ssufian/549294


## link to Github:

https://github.com/ssufian/Data_607


--------------------------------------------------------------------------------

\clearpage

## Problem Statement

For this assignment, you should take information from a relational database and migrate it to a NoSQL

database of your own choosing. 

--------------------------------------------------------------------------------

\clearpage

## Assignment 12 - MYSQL, MongoDB & R 

- For the relational database, I chose MySQL flights database

- For the NoSQL database, I use MongoDB 

--------------------------------------------------------------------------------

\clearpage

## Loading libraries

```{r,message=FALSE, warning=FALSE}
library("DT")
library(dplyr)
library(plotrix)
library(data.table)
library(RMySQL)
library(DBI)
library(rstudioapi)
require (ggplot2)
library(DBI)
```

## Extraction from MySQL Database

```{r,message=FALSE, warning=FALSE}
# connecting to flights database in MYSQL

# This line do not ask for user and password
con <- dbConnect(MySQL(), user ="root", password = "Chicke0n", port = 3306, dbname = "flights", host = "localhost")

# This line ask for user name and password
#con <- dbConnect(MySQL(), user = showPrompt('Username',"Username",default="root"), password = #askForPassword(), port = 3306, dbname = "flights",host = "localhost")

#extract all the tables into their own data frames.from flights database

```

## Converting all database extract to tables

```{r}
airlines <- dbReadTable(con, "airlines")
airports <- dbReadTable(con, "airports")
flights <- dbReadTable(con, "flights")
planes <- dbReadTable(con, "planes")
weather <- dbReadTable(con, "weather")


```

## Merging tables together

- Data Transformation

- View the all the related flights tables

```{r}

# show the tables retreived from MYSQL

View(airlines)

View(airports)

View(flights)

View(planes)

View(weather)
```

## Filtering only on early depature and early arrival flights

```{r}
data <- flights %>% 
    filter( arr_delay <0 | dep_delay <0) %>% # taking only early arrivals & departures
    merge(airlines) %>% 
    rowwise() %>% 
    mutate(carrier_name = gsub("[\r\n]", "", name))
# drop the columns with return carriage
drop <- c('name')
data1 <- data[,!(names(data) %in% drop)]

head(data1, n=1)    
    
```

## Left Join flights to airports

- based only the intersections

```{r}
drop <- c('faa')

airports1 <- airports %>% mutate(dest = faa)

airports1 <- airports1[,!(names(airports1) %in% drop)]

data2 <- data1 %>% 
    left_join(airports1)
```

## Left Join flights to planes & Weather

- based only the intersections


```{r}
data3 <- data2 %>% 
    inner_join(planes)

data_final <- data3 %>% # Getting only the intersections of all the various tables
    inner_join(weather)

```

--------------------------------------------------------------------------------

\clearpage

## Load Data into MongoDB/Get connection first

- Loading data into a MongoDB was difficult for me.  I spent 3 days trouble shooting it but finally it 

worked:  

(1) The key was to get MongoDB installed (whether via cloud based on an on-prem version)

(2) Followed these steps in installing MongoDB on Windows;

https://www.guru99.com/installation-configuration-mongodb.html

## Starting MongoDB server

(3) Watch the following youtube video;

https://www.youtube.com/watch?v=JBEKJflNV2g.  This video taught me how to complete this project step by

step.  The key was to have MongoDB server installed and running (if you have an on-prem version).  Once

that is done, you can then complete the last step of migrating your curated tables from Mysql to Mongodb

--------------------------------------------------------------------------------

\clearpage

- Inserting flights database that I had "filtered" to my liking

- Making sure all tables were inserted properly

--------------------------------------------------------------------------------

\clearpage

```{r}
library(mongolite)
#start server running before trying code
#connect to db
mdb = mongo(collection = "data_final", db = "flights")

mdb$insert(data_final)
#mdb$remove('{}')
mdb$count()
```

```{r}

results <- mdb$find('{}') %>% 
    group_by(carrier_name, name) %>% 
    
    top_n(-10,arr_delay) 
     

head(results,n=1)
```

## Database Test Run

- To test out the new database, Let’s say of the early "birds", we wanted to know Which airlines has the

best arrival time in the past year of 2013, i.e., the best of the best on-time performers?  

- Note: Here negative values means early arrival and departures; the more negative the values, the 

earlier the flight arrived

```{r}

#bar plot of Total raw ratings
ggplot(data=results, aes(x=carrier_name, y=arr_delay)) + geom_bar(stat="identity" )+
    geom_bar(colour="black", fill="#DD8888", width=.8, stat="identity" ) + 
    guides(fill=FALSE) +
    xlab("Airlines") + ylab("Early Arrival Scores") + coord_flip()+
    ggtitle("Airlines with Most Early Cummulative Arrival Occurances")

```

--------------------------------------------------------------------------------

\clearpage

Summary Results:

Since we filtered only on early arrivals and early depature flights, you are seeing only the best 

performance airlines in 2013.  

Of the early arrival airlines, United has the most number of cummalative early arrivals followed by 

Alaska Air and Jet Blue Airways

--------------------------------------------------------------------------------

\clearpage

Observation on the differences between traditional SQL and NOSQL:

- The major difference between a NoSQL and relational database is the schemaless architecture. This is

an advantage because it gives a lot of flexibility. It is nice to be able to add or drop data elements

as needs change.

- The main disadvantage is because of its "Anti SQL" nature, it is still very foreign to alot of data

engineers who grew up handling data structures the structured way. There's new syntax to be learn and 

that could be a hinderence for the old school data people.


--------------------------------------------------------------------------------

\clearpage



