
### title: "Data 607 - Project 2a"

author: "Sufian"

date: "9/30/2019"

output: html_document


Rpub link:

http://rpubs.com/ssufian/535540


Source:  United Nations Population Division
    
         Department of Economic and Social Affairs :  UN  International Migrant Stock

## Jai's list of Questions from his Discussion 5 post:

NOTE:  Migrant Data refers to Immigration into area of destination; area, region or countries

Jai posed these questions:  Summaries could be done by gender at any of the aggregation levels below

•	Gender ratios of migrant stock for each region of the world, for each income group, etc. 

•	Average gender ratios of world migrant stock. 

•	What is the variance across countries. 

•	Is there a trend across years for any of these sequences. 


## This project will focus only the following 2 regions straddling between high vs. low income:

- Developed vs. Less Developed Regions

- High vs. Low Income Countries

--------------------------------------------------------------------------------

\clearpage

## Loading libraries


```{r setup, include=FALSE}
# Load necessary libraries

library(janitor)
library(zoo)
library(tidyverse)
library(magrittr)
library(stringr)
library(DT)
library(tidyr)
library(dplyr)
```

## Loading Data from United Nations Reports 

```{r}
url <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/UN_MigrantStockTotal_2019%20(1).csv'

# Reading & Loading data
df <- read.csv(file = url ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE, stringsAsFactors = F, skip=13,header=F)

head(df)

#labeling columns 
new_name <- c("Sort","Region","Notes","Code","Data_type","1990.Total","1995.Total","2000.Total","2005.Total","2010.Total","2015.Total","2019.Total","1990.Male","1995.Male","2000.Male","2005.Male","2010.Male","2015.Male","2019.Male","1990.Female","1995.Female","2000.Female","2005.Female","2010.Female","2015.Female","2019.Female")


#Rename Columns
df <- df %>% 
     rename_at(vars(starts_with("V")), funs(gsub(.,"V",new_name))) 

head(df)

```

## Data Munging using Dplyr & TidyR


```{r}

df1 <- df

df1[df1==".."] <- "0"

# making dataset long format
df1 <- gather(df1,"year_types","n_years",6:26)
head(df1)

df2<-df1 %>% 
  mutate(n_years=str_replace_all(n_years,",","")) %<>% mutate_at(7, as.numeric)

head(df2)

#segregate total years into male and female years 

separate_DF <- df2 %>% separate(year_types, c("Year", "gender"))
head(separate_DF)

wide_DF <- separate_DF%>% spread(gender, n_years)
head(wide_DF)               
   
no_zero_DF_wide <- wide_DF %>% filter(Female != 0)

# Drop the unnecessary columns of the dataframe

no_zero_DF_wide <- select (no_zero_DF_wide,-c(Notes,Code,Data_type)) %>% mutate_at(3, as.integer) 
              
head(no_zero_DF_wide)

```

```{r }

no_zero_DF1 <- gather(no_zero_DF_wide, "gender","N_years",4:6)
                     
head(no_zero_DF1)                   
                    
```

## Migration Trends Studies (by different levels of economic & developed regions)

```{r}
require(ggthemes)

world_trend <- filter(no_zero_DF1, Region == 'WORLD') %>% 
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(world_trend)
#trendline plots of World Migrants
# Multiple line plot
ggplot(world_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("World - Migration Trendlines % terms: Men vs Women")+
  theme_excel()

```

--------------------------------------------------------------------------------

\clearpage

Observation 1a:

Total World Migrations by men & women have been rising from 1990 thru 2020 with men consistently higher than women

Men seems to be more mobile than women across time 

--------------------------------------------------------------------------------

\clearpage

## Drivers of World Migration Trends by gender (abs numbers)

```{r}
require(ggthemes)
library(ggplot2)
library(formattable)
#world migration over the years male vs. female
# Basic barplot - migration patterns over the years; men vs women

world <- no_zero_DF1 %>% filter(Region == "WORLD")
head(world)
 world %>% group_by(gender, Year) %>% 
    ggplot(aes(x=Year, y=N_years,fill = gender)) + 
    geom_bar(stat = "identity", position = "dodge") +
  xlab("Year") + ylab("Int'l Migrant Stock") +
  ggtitle("Drivers of World Migration Trends by Gender") + ylim(0, 271642105)+
   theme_excel()
```

--------------------------------------------------------------------------------

\clearpage

Observation 1b:

Total migratory growth mainly driven by Men in abs. terms across time; confirming the first chart

--------------------------------------------------------------------------------

\clearpage

## Variances of World Migration Trends by gender

```{r}
# Investigating variances of World Migration trends by Income & Development regions

#variance by male & female
world_trend_variance <- no_zero_DF1 %>% 
                     group_by(gender) %>% 
                      filter(gender == "Male" |gender == "Female" ) %>% 
filter(Region== "High-income countries" | Region=="Low-income countries"| Region=="More developed regions"| Region=="Less developed regions") %>% 
                mutate(std_dev = sd(N_years)) 

#Boxplot to show variances between regions and gender

p <- ggplot(world_trend_variance, aes(y=N_years,x=Region,fill=Region))+geom_boxplot()+
  ggtitle("Variances of World Migration by Gender from FY90 - FY20")+facet_grid( ~ gender)+
    theme_excel()
p <- p + theme(axis.text = element_text(size = 10,angle =45, hjust = 1))

p
```

--------------------------------------------------------------------------------

\clearpage

Observation 2a:

- Both High Income countries and more developed regions experienced greater immigration influxes based on their

medians with men seeing greater variances.  Not surprisingly, less developed regions and lower income countries saw

lower immigration based on their much lower medians and it also had lower spread in its

distributions


--------------------------------------------------------------------------------

\clearpage

###  Lets go underneath the regions and the income groupings to better understand the 

### underlying migration patterns of the 2 genders better

--------------------------------------------------------------------------------

\clearpage


```{r}
# Trendlines by Hi Income Group
Hi_Income_trend <- no_zero_DF1 %>% filter( Region == 'High-income countries') %>% 
              group_by(Year) %>% 
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
            
head(Hi_Income_trend) 

#trendline plots of Hi Income Migrants

ggplot(Hi_Income_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Hi Income Group - Migration Trendlines % terms: Men vs Women")+
  theme_excel()

# trendline plots of More Developed Regions Migrants
More_developed_trend  <- no_zero_DF1 %>% filter( Region == 'More developed regions') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(More_developed_trend) 
#trendline plots of More Developed Region Migrants

ggplot(More_developed_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("More Developed Region - Migration Trendlines % terms: Men vs Women")+
  theme_excel() 
#--------------------------------------------------------------------------------------------
# trendline plots of Low Income Migrants
Lo_Income_trend  <- no_zero_DF1%>% filter(Region == 'Low-income countries') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(Lo_Income_trend)
#trendline plots of low Income Migrants

ggplot(Lo_Income_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Low Income Group - Migration Trendlines % terms: Men vs Women")+
  theme_excel()            

# trendline plots of Less Developed Regions Migrants
Less_developed_trend  <- no_zero_DF1 %>%  filter(Region == 'Less developed regions') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(Less_developed_trend)
#trendline plots of Less Developed Migrants

ggplot(Less_developed_trend  , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Less Developed Region - Migration Trendlines % terms: Men vs Women")+
  theme_excel() 
```

--------------------------------------------------------------------------------

\clearpage

Observation 2b:

The various charts above were performed within each income countries and development regions to

see how each gender's(men vs. women) migatory patterns changed over time. 

- Hi Income countries:  Men is the dominant immigrant group relative to women but they diverged

after FY20 with men 

growing while women decreased

- Less Developed Region:  Simiarly, Men again is the dominant immigrant group with men

increasing post FTY20 while 

women decreased; similar divergent trendline as in Hi Income countries

- More Developed Region:  Surprisingly, Women is the dominant group here; trending at an

elevated level vs.men. The interesting fact was, both genders did not see any growth (flat

line) for almost 2 decades before women started to diverge and grew while surprisingly, men

decreased around FY2005.  It was until late 2018's that men rebounded slightly and the women

decreased simultaneously at the opposite end

- Low Income countries:  Perhaps the most interesting trends of all is this sub-group. Women

were trending higherearly FY20 and then criss-crossed with men in mid FY2005.  Post FY2005, men

decreased while women's numbers soared


--------------------------------------------------------------------------------

\clearpage

### Trying to understand the contradictory migration patterns  between more developed 

### regions vs. high income countries and less developed regions vs. low income countries?


(1) Let's group known developed nations and high income together next to known less developed

nations that are also low income. For example, G7 countries vs. 7 Less developed and Low Income

Countries, which were randomly selected from each hemisphere

(2) Then make a migratory trend comparision along gender lines to see if there were any 

discerning (clear) migration behavior between genders.


Note:  My hunch is that the way United Nations grouped the countries into various categories

may have cross-listed countries leading to this contradiction


--------------------------------------------------------------------------------

\clearpage


```{r}

#Segregating countries into poor (randomnly picked from each hemisphere) vs rich (G7 nations)

rich_countires <- no_zero_DF1 %>% filter(Region == 'United States of America'|Region ==  'Canada'|Region ==  'France'|Region == 'Germany'|Region ==  'Italy'|Region == 'United Kingdom'|Region == 'Japan') %>% 
              group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
  
  
poor_countries <- no_zero_DF1 %>% filter(Region == 'Albania'|Region =='Venezuela (Bolivarian Republic of)'|Region == 'Mexico'|Region == 'Honduras'|Region =='Syrian Arab Republic'|Region =='Egypt'
|Region =='Senegal') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))

poor_countries_men <- poor_countries %>% 
                      filter(gender == "Male" | gender == "Female") %>% 
                      group_by(gender) %>% 
                      summarise(sum(N_years)) 

rich_countries_men <- rich_countires%>% 
                      filter(gender == "Male" | gender == "Female") %>% 
                      group_by(gender) %>% 
                      summarise(sum(N_years))  

head(poor_countries)
head(rich_countires)
# bar plots of poor countries
ggplot(poor_countries, aes(x=Year, y=percent_migration_trends,fill = gender)) + 
    geom_bar(stat = "identity", position = "dodge") +
  xlab("Year") + ylab("Int'l Migrant Stock") +
  ggtitle("Drivers of World Migration in poor countries by Gender") +
   theme_excel()
# bar plots of G7 "rich" countries
ggplot(rich_countires, aes(x=Year, y=percent_migration_trends,fill = gender)) + 
    geom_bar(stat = "identity", position = "dodge") +
  xlab("Year") + ylab("Int'l Migrant Stock") +
  ggtitle("Drivers of World Migration in G7 countries by Gender") +
   theme_excel()

```

--------------------------------------------------------------------------------

\clearpage


Observation 3a:

By randomly picking known poor and less deveoped countries vs. known rich and developed 

countries such as the G7 nations, it was more visible which gender was more mobile (dominant).

It should be noted also that a more rigorous and robust way to select poor vs. rich countries

is via each countries' GDP or other socio-economic metrics.  This extra step was to quickly

answer the very confusing and contracdictory patterns that emerged in the  deep dive into

the constituents of the different groupings in Observation 2b

(1)  Visually from the Barcharts, Males were  more mobile over the last 3 decades  in poorer

nations while the opposite it true for richer nations that made up of the G7 nations; in fact

now that rich vs. poor countries were segregated distinctly to avoid cross-over

categorization, women actually were slightly more mobile 51% to 49% relative men in the rich

nations. While the opposite is true in the poorer countries.


--------------------------------------------------------------------------------

\clearpage

### One last question was in the poorer Nations, which country experienced the most

### number of migration?


The top 3 countries that experienced the largest migration flows were Venezuela, Syria and 

Mexico for the last 3 decades:  1990 to 2019.  This was not surprising as these 3 were conflict

nations; Venezuela had the most punishing economic collapse, the Syrians had a massive civil 

war while sadly enough, Mexico had the worst civilian violence stemming from the Narco trade


```{r}
most_migrant_countries <- no_zero_DF1 %>% filter(Region == 'Albania'|Region =='Venezuela (Bolivarian Republic of)'|Region == 'Mexico'|Region == 'Honduras'|Region =='Syrian Arab Republic'|Region =='Egypt'
|Region =='Senegal') %>% 
               group_by(Year) %>% 
               filter(gender=="Total") %>% 
                          arrange(desc(N_years))

most_migrant_countries 

ggplot(most_migrant_countries, aes(x=Year, y=N_years/sum(N_years),fill = Region)) + 
    geom_bar(stat = "identity", position = "dodge") +
  xlab("Year") + ylab(" Migrant Stock") +
  ggtitle("Driver countries of Migration in poorer nations") +
   theme_excel()
```

--------------------------------------------------------------------------------

\clearpage

## World migratory Variations over time 

```{r}

# comparing world migration variances over ALL Regions over time

world_migration_wide <- no_zero_DF_wide %>% 
                   group_by(Region, Year) %>% 
                   mutate(femalepct = Female/Total) %>% 
                   mutate(malepct = Male/Total) 
                  

head(world_migration_wide)
```
```{r}

dfhist <- world_migration_wide %>% 
         group_by(Year) #%>% 

# Overlaid histograms
pf <- ggplot(dfhist, aes(x=femalepct, color=Year)) +
  geom_histogram(fill="red", alpha=0.5, position="identity")+facet_grid(Year ~ .)
pf


pm <- ggplot(dfhist, aes(x=malepct, color=Year)) +
  geom_histogram(fill="yellow", alpha=0.5, position="identity")+facet_grid(Year ~ .)
pm

```

--------------------------------------------------------------------------------

\clearpage

Observation 3b:

Variance Analysis showed that both men and women Total World migratory distributions over time

exhibited normality.  

The only difference was women had a left-skew while men had a right skew.  Their central

tendencies were very similar as well.  The right skewness in men also confirmed the earlier box

plots of the hight income & high developed region groupings which was showing that men relative

to women had a higher spread due to longer positive tails.  

--------------------------------------------------------------------------------

\clearpage


## Summary Page

This short study had shown very interesting migatory behaviors between men and women over the

time periods:  1990 to 2019.  In general, men seem to be more mobile and was able to move into

higher income countries.  It was also shown that men were able to move more into the "less

developed regions" as well.  However, what is paradoxical, were the trends that showed women

were outpacing men in the "developed regions" and was also better in the low income

countries.  This was an ironic "finding" that deserved further investigation and analysis to

say the least.  Because these two sets of findings seems to be in contradiction.  The next

step of this study was to truly separated out the traditionally known rich nations relative to

to the poorer nations.  I randomly picked 7 poor nations from each hemisphere and compared it

to the G7 countries:

### The extra step show the following observations:

-  Once countries were unpacked, we clearly saw that women tend to be more mobile relative to 

men at 51% vs. 49%.  This statistics curiously was exactly the opposite in poorer countries

with men having the slight advantage.  

- It was also shown that within the poor countries; Venezuela, Syria and Mexico had the most

migration flow activities, most probably stemming from its internal socio-economic issues


--------------------------------------------------------------------------------

\clearpage




