---
title: "TidyVerse - a programming sample “vignette”"
author: "Sufian"
date: "11/5/2019"
output: html_document
---

Rpubs link: http://rpubs.com/ssufian/547141


--------------------------------------------------------------------------------

\clearpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Problem Statement:

In this assignment, we will practice collaborating around a code project with Github as a class.

Using several TidyVerse packages, and the "How Popular is Trump?"  dataset from

fivethirtyeight.com, I’m going to show a programming sample “vignette” that demonstrates how to use the

capabilities of ggplot2, dplyr, readr packages with this particular dataset.


```{r message=FALSE, warning=FALSE}

# Load
library("tidyverse")
library(ggplot2)
library(dplyr)

```

## Using readr to read data from a csv file (Github site)

--------------------------------------------------------------------------------

\clearpage


Part I:  Sufian's code and vignette

--------------------------------------------------------------------------------

\clearpage


```{r}
polls <- read_csv("https://raw.githubusercontent.com/ssufian/Data_607/master/approval_topline.csv")


polls1 <- read_csv('https://raw.githubusercontent.com/ssufian/Data_607/master/polling_data.csv')


```

##  Using ggplot2 to visualizae data; with pipe operation %>% from dplyr

```{r}

polls1 <- as_tibble(polls1)

```

## Bar Charts of Trump's approval/disapproval ratings by pollster types shown below:

```{r}
polls1 %>% ggplot(aes(x=pollster, y=adjusted_approve, fill=pollster))+
  geom_bar(stat = "identity", position = "dodge") + 
  guides(fill = FALSE) +
  ggtitle("Trump's approval ratings by Pollster types")+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

polls1 %>% ggplot(aes(x=pollster, y=adjusted_disapprove, fill=pollster))+
  geom_bar(stat = "identity", position = "dodge") + 
  guides(fill = FALSE) +
  ggtitle("Trump's disapproval ratings by Pollster types")+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## Trumps Approval/Dis-approval ratings by voter types:

```{r}
polls1 %>% ggplot(aes(x=subgroup, y=adjusted_approve, fill=subgroup))+
  geom_bar(stat = "identity", position = "dodge") + 
  guides(fill = FALSE) +
  ggtitle("Trump's approval ratings by voter types")+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

polls1 %>% ggplot(aes(x=subgroup, y=adjusted_disapprove, fill=subgroup))+
  geom_bar(stat = "identity", position = "dodge") + 
  guides(fill = FALSE) +
  ggtitle("Trump's disapproval ratings by voter types")+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

--------------------------------------------------------------------------------

\clearpage

Observation:

- Interesting to note that the voter type that really matters in the 3 sub groupings are adult

registered Voters; this group have an average higher adj. dis-approval ratings for Trump relatively

speaking.

- In parrellel, voters that have an adj. approval ratings for Trump has an overall lower ratings

relative to the dis-approvals among their peers.

--------------------------------------------------------------------------------

\clearpage


## Tidyverse Capability 1.

Description:

- Using filter we can select rows of the data frame matching conditions.

- filter capability tutorial


```{r}

filter(polls1, samplesize >800 & samplesize < 1992)

```

##  Tidyverse Capability 2.

Description:

- Using select we can keep the selected variables

- select capability tutorial

```{r}

df<- select(polls1, c("subgroup","pollster","approve", "disapprove"))
head(df)

```

--------------------------------------------------------------------------------

\clearpage

## Part II:  Sie Siong Wong's code and vignette that Sufian will extend, see below

--------------------------------------------------------------------------------

\clearpage

```{r}



```

