---
title: "Data 607 HW3 - RegEX"
author: "Sufian"
date: "9/9/2019"
output: html_document
---

Rpub link:

http://rpubs.com/ssufian/527284

GitHub link:




--------------------------------------------------------------------------------

\clearpage


## Question 3


```{r}

library(tidyverse)
library(stringr)

#copy the raw data
raw.data <-"555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert"
 
raw.data


```

The names extracted from the original is shown below:

```{r, echo=FALSE}
name <- unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))

name

```

```{r}
library(stringr)

#extract last names
last1 <- str_extract(name, '[[:alpha:]]{1,}\\,|\\b [[:alpha:]]{1,}')
name
last <- str_extract(last1, "[[:alpha:]]{1,}")

last

```

```{r}
#extract first names

first1 <- str_extract(name, '[[:alpha:]]{1,} |\\. [[:alpha:]]{1,}|\\, [[:alpha:]]{2,}')


first <- str_extract(first1, "[[:alpha:]]{1,}")

first1

first
```

```{r}

# prob 3 cont'd:  putting firstname  lastname convention in dataframe
df <- data.frame(first,last)

df
```

```{r}
# logical check for titles or not in dataframe

df$title <- str_detect(name, 'Rev.|Dr.')

df

```
```{r}
# logical check if there's middle name in dataframe
df$middlename <- str_detect(name, ' [[:alpha:]]{1,}\\.')

df
```
```{r}
# create a new column with first and last names in FINAL dataframe

df$first_last <- paste(df$first,df$last, sep=" ",collapse=NULL)
df
```

--------------------------------------------------------------------------------

\clearpage

## Question 4 and its associated 5 sub questions 

1. [0-9]+\\$

ans: 

This regex returns a string of numericals followed by $ sign 

```{r}
test <- "999$a###6-77$.2.2$a11$"

unlist(str_extract_all(test, '[0-9]+\\$'))

```
2. \\b[a-z]{1-4}\\b

ans: 

This regex returns 1-4 lower alpha characters  bounded at both ends with  a leading non-word and an ending non-word

```{r}
test2 <- "z-one3 55.tree !CRAZY%$-czdz"

unlist(str_extract_all(test2, '\\b[a-z]{1,4}\\b'))
```

3.  .*?\\.txt$

ans: 

This regex returns anything with as many repititions, 0 or more times that ends with a litteral of .txt

```{r}

test3 <- "911!889]]2.txt else.pdf----%&@.txt"
unlist(str_extract_all(test3, '.*?\\.txt$'))
```

4.  \\d{2}/\\d{2}/\\d{4

ans: This regex returns a date format type such 2 digit month, 2 digit days followed by 4 digit year:  two digits(usually months) followed by another two digits(days) and then four digits(years)


```{r}
test4 <- "09/11/2019 02/02/2002 1986/01/02"
unlist(str_extract_all(test4, '\\d{2}/\\d{2}/\\d{4}'))
```

5.  

<(.+?)>.+?</\\1>

ans:

This regex matches any <text > that ends with <same text>


```{r}

test5 <- "<oor><pi>jacki007<//ollaah>12ee22<li>once upon a time in 1999 </oor>"

unlist(str_extract_all(test5, '<(.+?)>.+?</\\1>'))


```

--------------------------------------------------------------------------------

\clearpage
## Question 9. Secret Message
The following code hides a secret message. Crack it with R and regular expressions. Hint: Some of the characters are more revealing than others! The code snippet is also available in the materials at www.r-datacollection.com.
```{r}

encrypted <- "clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0TanwoUwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigOd6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr"


## remove all nums, lowercase letters
cracked <- unlist(str_extract_all(encrypted, '[^[a-z]|[0-9]]'))

cracked

cracked1 <- paste0(cracked, collapse="")

cracked1

gsub(".", " ", cracked1,fixed=TRUE) 
```

