---
title: "Data 607 HW5"
author: "Sufian"
date: "9/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prob Statement

--------------------------------------------------------------------------------

\clearpage

(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. 

You’re encouraged to use a “wide” structure similar to how the information appears above, so that you can

practice tidying and transformations as described below. 

(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and

transform your data.   

(3) Perform analysis to compare the arrival delays for the two airlines. 

(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative

descriptions of your data cleanup work, analysis, and conclusions. 

--------------------------------------------------------------------------------

\clearpage

RPubs Link:

http://rpubs.com/ssufian/532710

Github Links:


https://github.com/ssufian/Data_607

--------------------------------------------------------------------------------

\clearpage

## Loading data and importing necessary packages

```{r}
library(janitor)
library(zoo)
library(tidyr)
library(dplyr)

#loading "wide" data from Github Repo


url <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/airlines.csv'

df1 <- read.delim(file =url, header=T , stringsAsFactors =FALSE, sep = ",")

df1 # checking the raw data sets

df1 <- df1 %>% mutate_all(na_if,"")%>%  
  do(na.locf(.)) %>% 
  distinct %>% # remove duplications after filling in empty lines with previous row
  rename(Airlines= X, arr_time = X.1) 


df1 <- gather(df1,"destination","N_flights",3:7)

df1


df2 <- df1 # to use as spread 

df1 <- df1 %>% 
  group_by(Airlines, arr_time) %>% 
  summarise(n_flights=sum(N_flights)) %>% 
  group_by(Airlines) %>% 
  mutate(percent_of_flights = n_flights/sum(n_flights))

# to get a dataframe in the format to compute the overall arrival performance of the airlines
df1 


#seperating delays and on-time into 2 columns
df2 <- spread(df2,arr_time, N_flights)

ctm2 <- df2 %>%
        clean_names() #using janitor library to remove spaces and put int underline in column name

ctm2

ddf3 <- ctm2 %>% 
    group_by(airlines,destination)%>% 
    mutate(percent_delayedflts = delayed/sum(delayed, on_time)) %>% 
    mutate(tot_flts = sum(delayed, on_time))
  
# per city performance grouped by airlines % destinations
ddf3 

ddf4 <- ddf3 %>% 
        group_by(airlines) %>% 
        mutate(total_flights = sum(tot_flts)) %>% 
        summarise(by_air= sum(tot_flts))

ddf4
```

--------------------------------------------------------------------------------

\clearpage

Summary of actions taken during the data munging process:

- After loading dataset, I took to make sure I have complete lines of data by filling in with NA and 

  replacing it with last observed data (LOCF) in order to fill out any missing rows
  
- I further went on to remove any duplications that may have arise due to the process of filing in blanks

- Mutating 2 new variables: % of delays, ranking. Renaming any columns necessary for further analysis

- I finally used the dplyr & Tidyr packages to get to the dataset needed to perform further quantative

   analysis

  (a) Making "wide" to "long" format and then utilize the groupby functions: airlines and arrival time
  
  (b) Perform aggregation 
  
  (c) mutating to create a new column; % of delays 

--------------------------------------------------------------------------------

\clearpage

## Data Analysis 1

Comparision of overall performance between the two airlines 

--------------------------------------------------------------------------------

\clearpage


```{r}
require(ggthemes)
library(ggplot2)

# Basic barplot - abs numbers
df1 %>% group_by(Airlines, arr_time) %>% 
  group_by(percent_of_flights) %>% 
  ggplot(aes(x=Airlines, y=n_flights, fill = arr_time)) +
  geom_bar(stat="identity", width = 0.5, position="dodge")+
  geom_text(aes(label=n_flights), vjust=1.6, color="black",position = position_dodge(0.45), size=3.5)+
  ggtitle("Number of Delay Arrivals by Airlines")+
  theme_economist()

```
```{r}
library(formattable)

# Basic barplot - percentage %
df1 %>% group_by(Airlines, arr_time) %>% 
  group_by(percent_of_flights) %>% 
  ggplot(aes(x=Airlines, y=percent_of_flights, fill = arr_time)) +
  geom_bar(stat="identity", width = 0.5, position="dodge")+
  geom_text(aes(label=percent(round(percent_of_flights,digits=2))), vjust=1.6, color="black",position =  position_dodge(0.45),     size=3.5)+
  ggtitle("% of Delay Arrivals by Airlines")+
  theme_economist()
```

--------------------------------------------------------------------------------

\clearpage

Observation 1:

Looking at the abs. values of on-tme and delays of AM West relative to Alaska side-by-side, it seems that

the higher no. of  delays in AM West is "small" relative to the number of on-time arrival of AM West,

which is twice the amount compared to Alaska's. However, A second look by percentage basis, showed that

overall, AM West performed better than Alaska Airways; 11% delays for AM West vs. 13% delays Alaska.

But is it?

--------------------------------------------------------------------------------

\clearpage

## Data Analysis 2

Comparision between arrival delays for the two airlines per destinations of the two airways to explain 

this discrepancy...


```{r}
# Basic barplot - arrival performance of the 2 airlines by destinations
ddf3 %>% 
  ggplot(aes(x=destination, y=percent_delayedflts, fill = airlines)) +
  geom_bar(stat="identity", width = 0.5,  position="dodge")+
  geom_text(aes(label=percent(percent_delayedflts)), vjust=0.1, color="black",position = position_dodge(0.55), size=3.00)+
  geom_text(aes(label=destination), vjust=1.4, color="black",position = position_dodge(0.55), size=3.25)+
  ggtitle("Per city Delays of the 2 airlines")+
  theme_excel()
```

```{r}
#loading data again to perform arrival comparasion BUT by destination
dfdest <- read.delim(file =url, header=T , stringsAsFactors =FALSE, sep = ",")

## Fill in empty rows with NA and Updating rows with earlier/previous rows values
dfdest <- dfdest %>% filter(complete.cases(" "))%>% 
  mutate_all(na_if,"")%>%  
  do(na.locf(.)) %>% 
  distinct%>% 
  rename(Airlines= X, arr_time = X.1) 

dfdest

#making it long format
dfdest1<- gather(dfdest,"destination","N_flights",3:7)

dfdest1

# this has another layer of groupby; adding destination
dfdest1 <- dfdest1 %>% 
  group_by(Airlines, arr_time, destination) %>% 
  summarise(n_flights=sum(N_flights)) %>% 
  group_by(Airlines) %>% 
  mutate(percent_of_flights = n_flights/sum(n_flights)) %>% 
  mutate(rank = rank(percent_of_flights)) %>% 
  filter(arr_time =="delayed")

dfdest1

```

--------------------------------------------------------------------------------

\clearpage

Checking and confirming the actual numbers of delays per city of the two airlines

--------------------------------------------------------------------------------

\clearpage

```{r}
# Basic barplot - Numbers of arrival delays per city
dfdest1 %>% group_by(Airlines, arr_time, destination) %>% 
  group_by(destination) %>% 
  ggplot(aes(x=destination, y=n_flights, fill = Airlines)) +
  geom_bar(stat="identity", width = 0.5, position="dodge")+
  geom_text(aes(label=n_flights), vjust=0.1, color="black",position = position_dodge(0.55), size=3.00)+
  geom_text(aes(label=destination), vjust=1.4, color="black",position = position_dodge(0.55), size=3.25)+
  ggtitle("No. of Delays per Service Cities")+
  theme_excel()
```

```{r}
# Basic barplot - Total flights taken by 2 airlines per each city
ddf3 %>% 
  ggplot(aes(x=destination, y=tot_flts, fill = airlines)) +
  geom_bar(stat="identity", width = 0.5,  position="dodge")+
  geom_text(aes(label=(tot_flts)), vjust=0.1, color="black",position = position_dodge(0.55), size=3.00)+
  geom_text(aes(label=destination), vjust=1.4, color="black",position = position_dodge(0.55), size=3.25)+
  ggtitle("Total flights taken by the airlines per city")+
  theme_excel()
```
```{r}

# Basic barplot - Total no. of flts taken by each airlines
ddf4 %>% 
  ggplot(aes(x=airlines, y=by_air, fill=airlines)) +
  geom_bar(stat="identity", width = 0.5,  position="dodge")+
  geom_text(aes(label=by_air), vjust=0.1, color="black",position = position_dodge(0.55), size=3.00)+
  geom_text(aes(label=airlines), vjust=1.4, color="black",position = position_dodge(0.55), size=3.25)+
  ggtitle("Total no. of flights taken by the each airlines")+
 
  theme_excel()
```

--------------------------------------------------------------------------------

\clearpage

Observation 2:

By all measures, the per-city performance of Alaska Airways is better than AM West (as measured by abs.

numbers and percentage basis).  The only exception was Seattle in which Alaska was worse than AM West.

The reason, the overall performance of AM West first appeared better was because they flew more flights

overall, and as such,  with a bigger denominator in terms of total flights, the  percentage delays 

per total flights taken resulted in a  smaller ratio than Alaska.  In fact, AM West flew twice as

much as Alaska Airways did.


--------------------------------------------------------------------------------

\clearpage

