
### title: "Data 607 - Project 2a"

author: "Sufian"

date: "9/30/2019"

output: html_document



--------------------------------------------------------------------------------

\clearpage

Rpub link:

http://rpubs.com/ssufian/535540

Github link:


--------------------------------------------------------------------------------

\clearpage


## Jai's list of Questions from his Discussion 5 post:

NOTE:  Migrant Data refers to Immigration into area of destination; area, region or countries

Jai posed these questions:  Summaries could be done by gender at any of the aggregation levels below

•	Gender ratios of migrant stock for each region of the world, for each income group, etc. 

•	Average gender ratios of world migrant stock. 

•	What is the variance across countries. 

•	Is there a trend across years for any of these sequences. 

## This project will focus only the following 2 regions straddling between high vs. low income:

- Developed vs. Less Developed Regions

- High vs. Low Income Countries

--------------------------------------------------------------------------------

\clearpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Data from United Nations Reports (stpred in my Github repository)


```{r }
# Load necessary libraries

library(janitor)
library(zoo)
library(tidyverse)
library(magrittr)
library(stringr)
library(DT)
library(tidyr)
library(dplyr)

url <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/UN_MigrantStockTotal_2019%20(1).csv'

# Reading & Loading data
df <- read.csv(file = url ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE, stringsAsFactors = F, skip=13,header=F)

head(df)

#labeling columns 
new_name <- c("Sort","Region","Notes","Code","Data_type","1990.Total","1995.Total","2000.Total","2005.Total","2010.Total","2015.Total","2019.Total","1990.Male","1995.Male","2000.Male","2005.Male","2010.Male","2015.Male","2019.Male","1990.Female","1995.Female","2000.Female","2005.Female","2010.Female","2015.Female","2019.Female")


#Rename Columns
df <- df %>% 
     rename_at(vars(starts_with("V")), funs(gsub(.,"V",new_name))) 

head(df)

```

## Data Munging using Dplyr & TidyR


```{r}

df1 <- df

df1[df1==".."] <- "0"

# making dataset long format
df1 <- gather(df1,"year_types","n_years",6:26)
head(df1)

df2<-df1 %>% 
  mutate(n_years=str_replace_all(n_years,",","")) %<>% mutate_at(7, as.numeric)

head(df2)

#segregate total years into male and female years 

separate_DF <- df2 %>% separate(year_types, c("Year", "gender"))
head(separate_DF)

wide_DF <- separate_DF%>% spread(gender, n_years)
head(wide_DF)               
   
no_zero_DF_wide <- wide_DF %>% filter(Female != 0)

# Drop the unnecessary columns of the dataframe

no_zero_DF_wide <- select (no_zero_DF_wide,-c(Notes,Code,Data_type)) %>% mutate_at(3, as.integer) 
              
head(no_zero_DF_wide)

```

```{r }

no_zero_DF1 <- gather(no_zero_DF_wide, "gender","N_years",4:6)
                     
head(no_zero_DF1)                   
                    
```

## Migration Trends Studies (by different level of economic and different developed regions)

```{r}
require(ggthemes)

world_trend <- filter(no_zero_DF1, Region == 'WORLD') %>% 
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(world_trend)
#trendline plots of World Migrants
# Multiple line plot
ggplot(world_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("World - Migration Trendlines % terms: Men vs Women")+
  theme_excel()


```

--------------------------------------------------------------------------------

\clearpage

Observation 1a:

Total World Migrations by men & women have been rising from 1990 thru 2020 with men consistently higher than women

Men seems to be more mobile than women across time 

--------------------------------------------------------------------------------

\clearpage

## Drivers of World Migration Trends by gender (abs numbers)

```{r}
require(ggthemes)
library(ggplot2)
library(formattable)
#world migration over the years male vs. female
# Basic barplot - migration patterns over the years; men vs women

world <- no_zero_DF1 %>% filter(Region == "WORLD")
world
 world %>% group_by(gender, Year) %>% 
    ggplot(aes(x=Year, y=N_years,fill = gender)) + 
    geom_bar(stat = "identity", position = "dodge") +
  xlab("Year") + ylab("Int'l Migrant Stock") +
  ggtitle("Drivers of World Migration Trends by Gender") + ylim(0, 271642105)+
   theme_excel()
```

--------------------------------------------------------------------------------

\clearpage

Observation 1b:

Total migratory growth mainly driven by Men in abs. terms across time; confirming the first chart

--------------------------------------------------------------------------------

\clearpage

## Drivers of World Migration Trends by gender

```{r}
# Investigating variances of World Migration trends by Income & Development regions

#variance by male & female
world_trend_variance <- no_zero_DF1 %>% 
                     group_by(gender) %>% 
                      filter(gender == "Male" |gender == "Female" ) %>% 
filter(Region== "High-income countries" | Region=="Low-income countries"| Region=="More developed regions"| Region=="Less developed regions") %>% 
                mutate(std_dev = sd(N_years)) 

head(world_trend_variance)

p <- ggplot(world_trend_variance, aes(y=N_years,x=Region,fill=Region))+geom_boxplot()+
  ggtitle("Variances of World Migration by Gender from FY90 - FY20")+facet_grid( ~ gender)+
    theme_excel()
p <- p + theme(axis.text = element_text(size = 10,angle =45, hjust = 1))

p
```

--------------------------------------------------------------------------------

\clearpage

Observation 2a:

- Both High Income countries and more developed regions experienced greater immigration influxes based on their

medians with men seeing greater variances.  Not surprisingly, less developed regions and lower income countries saw

lower immigration based on their much lower medians and it also had lower spread in its distributions


```{r}
# Trendlines by Hi Income Group
Hi_Income_trend <- filter(no_zero_DF1, Region == 'High-income countries') %>% 
              group_by(Year) %>% 
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
(Hi_Income_trend) 
#trendline plots of Hi Income Migrants

ggplot(Hi_Income_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Hi Income Group - Migration Trendlines % terms: Men vs Women")+
  theme_excel()

# trendline plots of More Developed Regions Migrants
More_developed_trend  <- filter(no_zero_DF1, Region == 'More developed regions') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(More_developed_trend) 
#trendline plots of More Developed Region Migrants

ggplot(More_developed_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("More Developed Region - Migration Trendlines % terms: Men vs Women")+
  theme_excel() 
#-------------------------------------------------------------------------------------------------------------------
# trendline plots of Low Income Migrants
Lo_Income_trend  <- filter(no_zero_DF1, Region == 'Low-income countries') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(Lo_Income_trend)
#trendline plots of low Income Migrants

ggplot(Lo_Income_trend , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Low Income Group - Migration Trendlines % terms: Men vs Women")+
  theme_excel()            

# trendline plots of Less Developed Regions Migrants
Less_developed_trend  <- filter(no_zero_DF1, Region == 'Less developed regions') %>% 
               group_by(Year) %>%
               filter(gender == "Male" | gender == "Female") %>% 
               mutate(percent_migration_trends = N_years/sum(N_years))
               
head(Less_developed_trend)
#trendline plots of Less Developed Migrants

ggplot(Less_developed_trend  , aes(x = Year, y = percent_migration_trends)) + 
  geom_line(aes(color = gender), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  ggtitle("Less Developed Region - Migration Trendlines % terms: Men vs Women")+
  theme_excel() 
```

--------------------------------------------------------------------------------

\clearpage

Observation 2b:

The various charts above were performed within each income countries and development regions to see how each

gender's(men vs. women) migatory patterns changed over time. 

- Hi Income countries:  Men is the dominant immigrant group relative to women but they diverged after FY20 with men 

growing while women decreased

- Less Developed Region:  Simiarly, Men again is the dominant immigrant group with men increasing post FTY20 while 

women decreased; similar divergent trendline as in Hi Income countries

- More Developed Region:  Surprisingly, Women is the dominant group here; trending at an elevated level vs.

men. The interesting fact was, both genders did not see any growth (flat line) for almost 2 decades before women

started to diverge and grew while surprisingly, men decreased around FY2005.  It was until late 2018's that men

rebounded slightly and the women decreased simultaneously at the opposite end

- Low Income countries:  Perhaps the most interesting trends of all is this sub-group. Women were trending higher

early FY20 and then criss-crossed with men in mid FY2005.  Post FY2005, men decreased while women's numbers soared


--------------------------------------------------------------------------------

\clearpage

## World migratory Variations over time 

```{r}

# comparing world migration variances over ALL Regions over time

world_migration_wide <- no_zero_DF_wide %>% 
                   group_by(Region, Year) %>% 
                   mutate(femalepct = Female/Total) %>% 
                   mutate(malepct = Male/Total) 
                  

head(world_migration_wide)
```
```{r}

dfhist <- world_migration_wide %>% 
         group_by(Year) #%>% 

# Overlaid histograms
pf <- ggplot(dfhist, aes(x=femalepct, color=Year)) +
  geom_histogram(fill="red", alpha=0.5, position="identity")+facet_grid(Year ~ .)
pf
pm <- ggplot(dfhist, aes(x=malepct, color=Year)) +
  geom_histogram(fill="yellow", alpha=0.5, position="identity")+facet_grid(Year ~ .)
pm

```

--------------------------------------------------------------------------------

\clearpage

Observation 3:

Variance Analysis showed that both men and women Total World migratory distributions over time were quite normal.  

The only difference was women had a left-skew while men had a right skew.  Their central tendencies were very 

similar as well.  The right skewness in men also confirmed the earlier box plots of the 2 regions with 2 income

groupings which was showing that men relative to women had a higher spread due to longer positivve tails.  


--------------------------------------------------------------------------------

\clearpage


## Summary

This short study had shown very interesting migatory behaviors between men and women over the time periods:  1990 to 
2019.  In general, men seem to be more mobile and was able to move into higher income countries.  It was also shown 

that men were able to move more into the "less developed regions" as well.  However, what is paradoxical, was the

trends that showed women were outpacing men in the "developed regions" and was also better in the low income

countries.  This is ironic and deserved further investigations and analysis to say the least.  Because these two

sets of findings seems to be in contradiction.  The next phase of this study should be:

-  What were the constituents of all these sub-groupings?  What made up High income vs. low income and what 

countries rolled up into "more developed regions" vs. "less developed regions".  

-  Once we unpack that, we than have to see if there were any cross-listing of countries in one category with the 

other category; meaning is a low income country considered less developed only or there were low income countries 

categorized as high developed regions.  In other words, all the underlying countries needs to be unpacked to ensure

there were no "double counting" which may have skewed the results above.  

At this point, the above study is just the begining, and further investigations are needed to be explored in before

any conclusions can be made about the general migratory behavior of the genders over time.


--------------------------------------------------------------------------------

\clearpage




