
title: "Data 607 - Project 1"

author: "Sufian"

date: "9/18/2019"

output: html_document

Rpubs Links:
  
http://www.rpubs.com/ssufian/530973

GitHub Links:
  
https://github.com/ssufian/Data_607


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```

In this project, the task is to transform a text file with chess tournament results into a clean tidy data

format and store the results as a .CSV file with the following information for all of the players:

Player’s Name

Player’s State

Total Number of Points

Player’s Pre-Rating

Average Pre Chess Rating of Opponents

## Load the Tournament Data


```{r}
# Load necessary libraries
library(stringr)
library(DT)
library(tidyr)
library(dplyr)


url <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/tournamentinfo.txt'
mydata <- read.table(file =url, header=F , stringsAsFactors =FALSE, sep = ",")

mydata[1:10,]
         
```

## Replace dash lines with empty spaces 


```{r}
library (stringr)
newdata <- data.frame(str_replace_all(mydata$V1,"-",""))
newdata[1:10,]

```

## Removing empty spaces
```{r}
#take only populated rows that do not contain emptly lines
newdata <- as.data.frame(newdata[!apply(newdata == "", 1, all),])

newdata[1:10,]
```
```{r}

odd <- newdata[seq(1,nrow(newdata),2),] # only odd rows
even <- newdata[seq(2,nrow(newdata),2),] # only even rows

head(odd) #check out my odd rows

head(even) #check out my even rows

```

## Segregating my even and odd rows, checking it out

Merging rows into columns into a temp table; called new_table

```{r}
#Merging the odd & the even rows into a newly created empty place holder table

new_table <- data.frame(c())
new_table <- data.frame(paste(odd,even))

new_table[1:5,]
```

## Merge the even and odd rows in columns

Creating headers in the merge table while eliminating the pipes (|) deliminators and deleting first rows 

since the new headers were created; like starting from a clean slate of a new table

```{r}
# Create headers
Headers <- c("Pair","Player Name","Total","Round 1","Round 2","Round 3", "Round 4","Round 5","Round 6","Round 7","State","USCF ID / Rtg (Pre->Post)","Pts","1","2","3","4","5","6","7","Avg Pre Rating")

# Eliminating | and Separating into columns

df <- separate(data = new_table, col=paste.odd..even., into = Headers, sep = "\\|")

# HOusekeeping: Delete row containing all columns names
df <- df[-c(1), ]

df[1:5,]

```
```{r}
# Extracting numerical values from "USCF ID / Rtg (Pre->Post)" column for calculating averages later

temp <- str_extract_all(df$`USCF ID / Rtg (Pre->Post)`,"\\b\\d{1,}")
temp <- data.frame(as.character(temp)) #making into dataframe
head(temp)

# breaking out the columns; extracting the 3 numericals

# Split the data frame: 1 column -> 3 different columns

temp <- separate(data = temp, col = as.character.temp., into = c("col1","col2","col3"), sep = ",")

head(temp)

```

```{r}
# Temporary column vectors
ID <- str_extract_all(temp$col1,"[[:digit:]]{1,}")
Pre_rate <- str_extract_all(temp$col2,"[[:digit:]]{1,}")
Post_rate <- str_extract_all(temp$col3,"[[:digit:]]{1,}")

#converting ID, pre rating & post rating columns into numerics for calculation purposes
df$`USCF ID` <- as.numeric(ID)
df$`Pre Rating` <- as.numeric(Pre_rate)
df$`Post Rating` <- as.numeric(Post_rate)

head(df)

```

## Subsetting the temp table into smaller one and taking only the relevant columns

```{r}
df_new <- subset(df, select = c(1,2,11,22,3,23,24,21))
head(df_new)

class(head(df_new$'Pre Rating')) # check class type

class(head(df_new$`Avg Pre Rating`)) # check class type

# converting characters in Avg. Pre Rating columns to numeric
df_new$`Avg Pre Rating` <- 0

head(df_new)

df_new[39,] #double checking on a single Player
```

## Extricating numericals from opponents to compute its avg scores and create a Opponents only table

```{r}
# Extracting numerical values from rounds played (opponents against)
opp1 <- data.frame(as.numeric(str_extract_all(df$`Round 1`,"[[:digit:]]{1,}")))
opp2 <- data.frame(as.numeric(str_extract_all(df$`Round 2`,"[[:digit:]]{1,}")))
opp3 <- data.frame(as.numeric(str_extract_all(df$`Round 3`,"[[:digit:]]{1,}")))
opp4 <- data.frame(as.numeric(str_extract_all(df$`Round 4`,"[[:digit:]]{1,}")))
opp5 <- data.frame(as.numeric(str_extract_all(df$`Round 5`,"[[:digit:]]{1,}")))
opp6 <- data.frame(as.numeric(str_extract_all(df$`Round 6`,"[[:digit:]]{1,}")))
opp7 <- data.frame(as.numeric(str_extract_all(df$`Round 7`,"[[:digit:]]{1,}")))

#  build an Opponents data frame
opponents <- cbind(df$Pair, df$`Player Name`, opp1, opp2, opp3, opp4, opp5, opp6, opp7)
names(opponents) <- c("Pair","Player Name","Opp 1","Opp 2","Opp 3","Opp 4","Opp 5","Opp 6","Opp 7")

head(opponents) # Opponents only Dataframe

```

## Continuing with the Opponents only table, calculate how many games each of them had played 

```{r}
# Detecting any NA replaced and by 0.
opponents[is.na(opponents)] <- 0

head(opponents)
```
```{r}

opponents[is.na(opponents)] <- 0 # NA replacement by zero
count = 0
# Finding number of games played
for(i in 1:nrow(opponents)){
    for(j in 3:9){
    if(opponents[i,j]>0){
      count = count + 1
    }
    opponents$No_of_Games_Played[i] <- sum(count)
    
    }
  count = 0
}

head(opponents)

```

## Matching up players (via Pairing column) with its opponents played and compute the pre rating averages

```{r}

#Resetting Avg Pre Rating column to zero before calculating their averages
df_new$`Avg Pre Rating` <- 0

# Compute Average Pre-Rating for each player
for (k in 3:9){
  for (j in 1:nrow(df_new)){
    for (i in 1:nrow(df_new)){
      if (as.numeric(opponents[j,k]) == as.numeric(df_new$Pair[i])){
        df_new$`Avg Pre Rating`[j] <- as.numeric(df_new$`Avg Pre Rating`[j]) + as.numeric(df_new$`Pre Rating`[i])
      }
    }
  }
}

# Each player's average based on the number of played games
df_new$`Avg Pre Rating`  <- round(as.numeric(df_new$`Avg Pre Rating`) / opponents$No_of_Games_Played,0)

df_new[1:10,]
```

## Data Analysis via ggplotings 

```{r}
library(ggplot2)
library(plotly)
require(ggthemes)
# Scatter plot to Check if there's any reasonable trend?

p1 <- ggplot(df_new, aes(df_new$'Pre Rating', y = df_new$'Post Rating')) + geom_point(aes(color=State))+theme_economist()

p1 <- ggplotly(p1)

p1

fit <- lm(df_new$'Avg Pre Rating' ~ df_new$'Pre Rating', data = df_new)

p3 <- ggplot(df_new, aes(df_new$'Pre Rating', y = df_new$'Avg Pre Rating')) + 
  geom_point(aes(color=df_new$Total)) + theme_economist() +
  stat_smooth(method = "lm", col = "red") +
  
    ggtitle(paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),";
                     Intercept =",signif(fit$coef[[1]],5 ),";
                     Slope =",signif(fit$coef[[2]], 5),";
                     P =",signif(summary(fit)$coef[2,4], 5)))
  
  

p3 <- ggplotly(p3)

p3

```

## Observations from Scatterplots:

(1)  By States:  There appears to have a positive correlation between player's pre tournament ratings vs. 

     opponents' pre tournament ratings.  By states, they appear "matched" up pretty well in terms of
     
     skills.  This might have been to prevent lop-sided matches whereby, very good players are playing
     
     against less-skilled players.
     
(2)  The 2nd chart showed a comparison between the player’s own pre-rating vs. their opponents’ average

     pre-rating. The color corresponds to the total number of points scored in the tournament.
    
     All observations above the red diagonal line represent players had a higher pre-rating than
     
     the average of their opponents, and the points below the line is the reverse.

Conclusion:

    While chart 1 showed a "well-distributed" match ups of players across states, which is good for the
    
    game and the fans especially, chart 2 tells a different story;  Chart 2 gives suggest that its almost
    
    half and half whereby highly-ranked players were matched up against lower- ranked players (based on
    
    pre tournament rankings).  This is visually discerning by the Red regression line which is almost
    
    almost flat as evidenced by slope = 0.128.
```{r}
#histogram to confirm that first scatter plot (chart 1) was misleading

hist(df_new$`Avg Pre Rating`, main="Avg Opponents Pre Tournament Ratings", xlab="opponents' Avg pre rating")

hist(df_new$`Pre Rating`, main="Players' Own Pre Tournament Ratings", xlab="Players' Pre rating")

hist(df_new$`Post Rating`, main="Players' Own Post Tournament Ratings", xlab="Players' Post rating")
```

## Observations from Basic Histograms:

(1)  First 2 histograms of opponnents avg pre-tournament rankings vs. the players' own ranking showed

     a normal distribution of  their opponents' average ranking of around 1350.  However, the players' 
     
     own pre-tournament rankings was left-skewed with a higher ratings of around 1500.  These addtional 
     
     charts showed the likely hood of player mis-matches in terms of skill sets
     
```{r}
ggplot(df_new, aes(df$'State',df$`Pre Rating`)) +geom_boxplot()

ggplot(df_new, aes(df$'State',df$`Post Rating`)) +geom_boxplot()
```

## Observations of distribution of ranking scores from Box plots:

(1) By States again:  Except for Ohio which had only 1 entry (therefore no distribution), both Michigan

    and Ontario players' had higher median pre-tournment rankings vs. post tournment rankings.  Also, they
    
    had a smaller pre-tournament spread while not only their post-tournament spread became bigger, their
    
    medium ranking decreased. This showed that this tournament had an adverse effect on some players more 
    
    than others which could indirectly be explained by chart 2 (scatter plot above) which showed half of
    
    the players were matched up against players that were ranked higher than themselves.

```{r}

#Creating a data table 
datatable(df_new, rownames=FALSE)

```

```{r}
#Writing to CSV file
write.table(df_new,file="Sufian_Suwarman_Project1.csv",row.names=FALSE,col.names=TRUE,sep=",")
```


