---
title: "Data 607 Project 2c"
author: "Sufian"
date: "10/5/2019"
output: html_document
---


Rpub Link:

http://rpubs.com/ssufian/536313


GitHub Link:


https://github.com/ssufian/Data_607


--------------------------------------------------------------------------------

\clearpage


### One of the many Headlines about the Millennials:

### "Why Millennials Are The Most Screwed Generation Of All Time" 

     editorial piece by Veronica Walsingham in Ranker.com
     https://www.ranker.com/list/millennials-most-screwed-generation/veronica-walsingham
     
### (a) Data source: Federal Reserve Bank of New York Consumer Credit Panel / Equifax

### (b) Data source: Bureau of Labor Statistics - https://data.bls.gov/PDQWeb/ln


### Is the editorial headline about Millennails true? 

Part  1:

Let's explore and start by looking at their debt obligations relative to other generational groups, see below

(a) Baby Boomers: Born 1946-1964 (54-72 years old) 

(b) Generation X: Born 1965-1980 (38-53 years old) 

### Millennials: Born 1981-1996 (22-37 years old) 

(c) Generation Z: Born 1997-Present (0-21 years old)

This exercise will focus on two forms of debt outstanding:  Student loans and Credit Card loans


Part  2:

The 2nd phase is to look at their Job prospects thru the lens of economic growth indicators since the 

great recession of 2008;  GDP of economy and unemployement rate and tuition rates over time 


--------------------------------------------------------------------------------

\clearpage

### The objective of this exercise is to answer: Are the Millennials really screwed?

 - Let's start by looking at both ends of the economic spectrum facing this generation and then when all the facts are 

in, do we agree or disagree with the statement above.  Lets find out....?


--------------------------------------------------------------------------------

\clearpage



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(ggthemes)
library(tidyverse)
library(magrittr)
library(stringr)
library(tidyr)
library(dplyr)
library(lubridate)

```

## Loading Data, deleting unwanted columns


```{r }

url1 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/No_Student_Loan_by_Age.csv'

url2 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/Student_Loan_by_Age.csv'

# Reading & Loading data
df1 <- read.csv(file = url1 ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=3, stringsAsFactors = F,header=T)

df2 <- read.csv(file = url2 ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=3, stringsAsFactors = F,header=T)

df1 <- df1 %>% select(-c(X.1:X.11)) # Number of Student loan borrowers

df2 <- df2 %>% select(-c(X.1:X.142)) # Dollar amount in $millions of Student loans

```

## Data wrangling via Dplyr:  renaming pertinent columns


```{r }

oldnames = c("X30.39","X40.49","X50.59","X60.")
newnames = c("under30_39","under40_49","under50_59","over60")

dfclean1 <- df1 %>% 
  rename(Year=X) %>% 
  rename_at(vars(oldnames), ~ newnames) 

dfclean2 <- df2 %>% 
  rename(Year=X) %>% 
  rename_at(vars(oldnames), ~ newnames) 
head(dfclean1)
head(dfclean2)




```

### Using Tidyr to make Wide to Long formatted dataframes

```{r}
dfnew1 <- gather(dfclean1,"Age_group","n_borrower",2:6)

dfnew2 <- gather(dfclean2,"Age_group","amount_borrowed",2:6)
         
head(dfnew1)
head(dfnew2)
```

--------------------------------------------------------------------------------

\clearpage


### Looking at their Student Loans


--------------------------------------------------------------------------------

\clearpage


```{r}
pn <- ggplot(data = dfnew1, aes(x = Year, y = n_borrower, fill = Age_group)) + 
    geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Total Number (millions) of Student loan borrower by age group")+ 
  theme_excel()
pn


pd <- ggplot(data = dfnew2, aes(x = Year, y = amount_borrowed, fill = Age_group)) + 
    geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Total Amount ($Billions) of Student loan borrower by age group")+
  theme_excel()
pd
```

### Observation 1:

Since the recession, both Millenials and Gen Z had been the highest number of borrowers and the highest by dollar values

as well.  While Gen Z's borrowing rate started to plateu out by mid 2012, the Millenials' student debt rate kept growing.


--------------------------------------------------------------------------------

\clearpage

```{r}

url3 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/tot_debt_by_age.csv'

url4 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/Debt%20Share%20by%20Product%20Type%20and%20Age.csv'

# Reading & Loading data
df3 <- read.csv(file = url3 ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=3, stringsAsFactors = F,header=T)

df4 <- read.csv(file = url4 ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=2, stringsAsFactors = F,header=T)

df3 <- df3 %>% select(-c(X:X.13)) # Number of Student loan borrowers

df4 <- df4 %>% select(-c(X.1:X.11)) # Dollar amount in $millions of Student loans


```


--------------------------------------------------------------------------------

\clearpage


### Now looking at Total debt by age group and see how Millenials fare?

--------------------------------------------------------------------------------

\clearpage

```{r}

df3new <-df3 
 
oldnames = c("X18.29","X30.39","X40.49","X50.59","X60.69","X70.")
newnames = c("under18_29","under30_39","under40_49","under50_59","under60_69","over70")

dfclean3 <- df3new  %>%
  slice(1:(n()-2)) %>% 
  rename_at(vars(oldnames), ~ newnames) 

dfclean4 <- df4  %>% 
  rename_at(vars(oldnames), ~ newnames) %>% 
  slice(1:6) %>% 
  rename(credit_type=X)




dfclean3 <- dfclean3 #%<>% 
        # mutate(year= as.Date(year, format= "%d.%m.%Y"))
         
         
         
head(dfclean3)         
(dfclean4)


```


### Use Tidyr to make Wide to Long format dataframe again


```{r}

dfnew3 <- gather(dfclean3,"Age_group","tot_debt",2:7)

dfnew4 <- gather(dfclean4,"Age_group","amount_borrowed",2:7)


         
head(dfnew3)
head(dfnew4)

         
```

### Plots of Total debt by Age groups & by Product Types and Age groups


```{r}
library(basicTrendline)


p3 <- ggplot(dfnew3, aes(x=quarter,y=tot_debt,fill=Age_group)) + geom_bar(stat = "identity")+
      xlab("Year") + ylab("Amount Owed in $ Trillions") +
      ggtitle("Total Debt Balance by Age") +  
      theme_excel()
             
p3 <- p3+theme(axis.text = element_text(size = 10,angle =90, hjust = 1))+theme(legend.position="top")

p3


# Multiple line plot
dfnewline <- dfnew3
p3b <- ggplot(dfnewline , aes(x = quarter, y = tot_debt)) + 
  xlab("Year") + ylab("Amount Owed in $ Trillions")+
  ggtitle("Total Debt trendlines by Age") +
  geom_line(aes(color = Age_group), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800","red","azure","yellow","black")) +
   theme_excel()

p3b <- p3b+theme(axis.text = element_text(size = 10,angle =90, hjust = 1))+theme(legend.position="top")


p3b



```

```{r}
dfnew4 <-dfnew4 %>% 
         group_by(Age_group)

p4 <- ggplot(dfnew4, aes(x=Age_group,y=amount_borrowed,fill=credit_type)) + geom_bar(stat = "identity")+
      xlab("Age Groups") + ylab("Amount Owed in $Trillions") +
      ggtitle("Debt Share by Product Type and Age (2019 Q2)") +  
      theme_excel()
             
p4 <-p4+theme(axis.text = element_text(size = 10,angle =0, hjust = 0.5))+theme(legend.position="top")

p4

```


### Observation 2:

(1) The total amount of debt carried by Millenials is only 2nd after Gen X and Baby Boomers.  While all age groups saw a 

spike in debt load at the peak of the debt bubble in 2006, debt load rate of Millenials kept increasing.  Also worrying

was the increasing rates of all age groups relative to Millenials.  The Millenials is the youngest age group to

allow its debt load to increased while surprisingly enough, Gen Z which was the youngest to have kept its debt load

relatively flat post the great recession of 2006.


(2)  The latest breakdown in debt load as of Q2 2019 showed the following:

-  Millenials was the 2nd most indebted age group after Gen X with the bulk in mortgages followed by Student loans, Auto

and Credit Card debts

- Perhaps the bright spot was that Gen Z is the lowest debt carrying age group across all debt products.


--------------------------------------------------------------------------------

\clearpage

```{r}

url5 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/Tuition_Roomboard.csv'


# Reading & Loading data
df5 <- read.csv(file = url5 ,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=3, stringsAsFactors = F,header=F)

oldlist <- paste("V.", 1:32,sep="", collapse = NULL)
newname <- c("institution","1986","1987","1988","1989","1990","1991","1992","1993","1994","1995","1996","1997","1998","1999",	"2000","2001","2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013","2014","2015","2016",	"2017","2018")


df5short <- df5 %>% select(-c(V35:V37)) %>% 
            slice(14:15)
head(df5short)

#Rename Columns
df5clean <- df5short %>% 
     rename_at(vars(starts_with("V")), funs(gsub(.,"V",newname))) 


head(df5clean)

# making dataset long format
df5cleaner<- gather(df5clean,"years","tuition_fees",2:34)
df5cleaner <- df5cleaner %>% 
  mutate(tuition_fees=str_replace_all(tuition_fees,",",""))%<>% 
  mutate_at(2, as.numeric)

# replace $ with blank "" in the df$payment column.  and coerce that result to numeric
df5cleaner$tuition_fees = as.numeric(gsub("\\$", "", df5cleaner$tuition_fees ))


head(df5cleaner)

```

```{r}
# Multiple line plot:  2yr and 4 yr public schools
#dfnewline <- dfnew3

p5 <- ggplot(df5cleaner , aes(x = years, y = tuition_fees)) + 
  xlab("Year") + ylab("Tuition fees + Room & Board")+
  ggtitle("Avg. Tuition and Fees plus Room & Board (Unweighted) in Current Dollars") +
  geom_line(aes(color = institution), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
   theme_excel()

p5 <- p5+theme(axis.text = element_text(size = 11,angle =0, hjust = 1))+theme(legend.position="top")


p5

```


--------------------------------------------------------------------------------

\clearpage

Observation 3:

The average 2 and 4 year public instituitions (excluding private entities) in current dollars have been growing with 

52%  increase just in the last decade!  This shows you the new younger generations of Gen Z and Millennials are NOT

getting any breaks if they wanted to or are currently pursuing any forms of post-secondary education 


--------------------------------------------------------------------------------

\clearpage


###  So far, we have focus on the costs or liability side of the ledger facing this Millennial generation.

Let's turn our focus to the revenue side of the equation:  GDP of US economy and Unemployment rate of these generations


```{r}

url6 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/BLS_Unemployment%20SeriesReport-20191006171232_c15ff5.csv'


# Reading & Loading data
df6 <- read.csv(file = url6,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=12, stringsAsFactors = F,header=F)


df6short <- df6 %>% select(-c(V2:V5,V7:V11,V13:V17,V19:V23)) 
 

#labeling columns "
new_name <- c("year","under18_21","under25_34","under35_44","under45_54")


df6shortnew <- df6short %>% 
          rename(year="V1","under18_21"='V6', "under25_34"="V12","under35_44"="V18","under45_54"="V24")
         

head(df6shortnew )
           
```

```{r}
# making dataset long format
df6cleaner<- gather(df6shortnew ,"Age_group","unemployment_rate",2:5)

head(df6cleaner)

```

```{r}
# Multiple line plot:  Unemployment rate of various age groups


df6cleaner <- df6cleaner %>% 
              group_by(year)

p6 <- ggplot(df6cleaner , aes(x = year, y = unemployment_rate)) + 
  xlab("Year") + ylab("% Unemployment Rate ")+
  ggtitle("Umemployment rate; various age groups") +
  geom_line(aes(color = Age_group), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800","black","red")) +
   theme_excel()

p6 <- p6+theme(axis.text = element_text(size = 11,angle =0, hjust = 1))+theme(legend.position="top")


p6

```

--------------------------------------------------------------------------------

\clearpage


Observation 4:

The good news is unemployment rate of the Millennials has been trending downward since the great recession, its currently

less than 5%.  By the same token, Gen X and beyond are also seeing very good unemployment numbers; all under 5%. Of note,

the older you are, the better off you are as measured by unemployment rate.  However, the worrying sign is Gen Z is

still seeing some elevated numbers, hovering around 12%.  But over all, all groups are seeing a down shift which is a

 of today's general economic strength, which is a topic we will delve into next to finish off this discussion.

--------------------------------------------------------------------------------

\clearpage


### The GDP story since the Great Recession


```{r}


url7 <- 'https://raw.githubusercontent.com/ssufian/Data_607/master/GDP.csv'


# Reading & Loading data
df7 <- read.csv(file = url7,sep = ",", na.strings = c("NA", " ", ""), strip.white = TRUE,skip=3, stringsAsFactors = F,header=T)


df7short <- df7 %>% select(-c(Country.Code,Indicator.Name,Indicator.Code,X1960)) 
 

#labeling columns "


df7shortnew <- df7short %>% 
            rename_at(vars(starts_with("X")), funs(str_replace(., "X", "")))
         

```

```{r}

# making dataset long format
df7cleaner<- gather(df7shortnew,"year","N_GDP",2:59) %>% 
             filter(Country.Name == 'United States')%<>% 
             mutate_at(2, as.numeric)

head(df7cleaner)

```

```{r}
# Trend line plot:  US GDP since 1961


df7cleaner <- df7cleaner %>% 
              group_by(year) %>% 
              filter(year >=2000)

p7 <- ggplot(df7cleaner , aes(x = year, y = N_GDP) )+ 
  xlab("Year") + ylab("% GDP ")+
  ggtitle("GDP growth rate since Yr 2000") +
   geom_line(aes(color = Country.Name), size = 1.3) +
  scale_color_manual(values = c("yellow")) +
   theme_excel()

p7 <- p7+theme(axis.text = element_text(size = 11,angle =0, hjust = 1))+theme(legend.position="top")


p7
```

--------------------------------------------------------------------------------

\clearpage

Observation 5:

Since the Great Recession of 2008, the US economy has been growing and it wasn't till 2013 there after, that the economy

clawed its way back to pre-recession state.  It has been averaging around 2% growth the last 6 years or so and in 2017,

the slope turned upwards.  This upward trend is one of the biggest reason, unemployment rate in observation 4 has gone 

down.  Good economic growth has reduced unemployment rate across all age groups and Millenials' downward unemployement

rate  is part of that "feel good"  story. In fact, this is the classic proverbial phrase; "high tide lifts all boats".


--------------------------------------------------------------------------------

\clearpage


Summary:

The discussion started out with a blaring editorial headline: 

### "Why Millennials Are The Most Screwed Generation Of All Time" 

But after looking at the "check book" of the Millennial generation from both sides;  Liabilities vs. Revenue (GDP &

Unemployment rate), we are now able to answer this question without biases and anecdotes.  These are the facts:

-  Yes, Millennials are much in debt with the highest dollar student loan amount across all age groups

-  Yes, Millennials, along with Gen Z are the highest number of borrowers of student loan across all age groups

-  Yes, Millennials, along with all the other age groups are still incurring more total debt post recession

-  Yes, Public 2 & 4 year college tuition plus room & board fees are increasing with a whopping 52% increase in just 

   the last decade
   
-  Yes, Unemployment rate across all age groups is decreasing along with a positive GDP growth in the last 6 years


With all these statistics in the backdrop, Its an open secret that we are a nation of debtors.  Millennials is just one

of the many. Millennials are certainly up against a lot of challenges with a huge debt overhang.  But Millennials are NOT

the only generation facing this dilemma.  In fact, all the generations across ages are facing their own challenges with

debts on their backs.  The question then becomes how much debt becomes deterimental to personal growth.  That is, if one

was to take on more debt, will there be enough earnings derived from a better economic future to offset this debt burden?

Since all age groups are "screwed" at a certain level because of their debt load, the real issue is then will there be

enough of a career path to pay off all these debts.  Maybe the better headline should read something like this:  

### "Are there enough sustainable economic growth to help our generation of debtors?"




